---
title: "Q4 code tips sheet"
output: pdf_document
---

# Accessing values in a regression

Here are some codes you can use to access the value of the coefficients after running a regression. We first load the `iris` (a type of flower) dataset and run a regression to determine the factors that affect the length of an iris petal. 

```{r}
# load packages
pacman::p_load(dplyr, tidyverse)

# load data
data(iris)

# look at what the data is like
head(iris)

# run a linear model
model1 <- lm(Petal.Width ~ Sepal.Length + Sepal.Width + Petal.Length, data = iris)
summary(model1)

# get the intercept
model1$coefficients[1]

# get coefficient of petal.length
model1$coefficients['Petal.Length']

# to get the coefficients of sepal.length and sepal.width (second and third variables in the reg)
model1$coefficients[2:3]
```
# Pivot Tables in R

You can use `group_by()` and `summarise()` functions to replicate Pivot Tables in R. The `iris` dataset contains 3 species (`table(iris$Species)`) and we want to see the average petal length and width of each species.

```{r, message = F}
petal_pivot <- iris %>%
  group_by(Species) %>%
  summarise(avg_width = mean(Petal.Width),
           avg_length = mean(Petal.Length))

petal_pivot
```

The `petal_pivot` object is a tibble, which is a special type of a dataframe (if you type `class(petal_pivot)`, it will say `tbl_df`). To access values in a tibble, you will have to use double brackets instead of single brackets like you would with a data frame. Alternatively, you can call use the variable name with the `$` operator to look at the `avg_length` column and use single brackets to extract the value for the setosa row. 

```{r}
petal_pivot[[1, 3]]
petal_pivot$avg_length[petal_pivot$Species == "setosa"]
```


# Creating vectors using `rep()` and `seq()`

To replicate elements of vectors, such as your July futures price, you can use the `rep()` function (documentation [here](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/rep)). The syntax is `rep(value, number_of_times)`. The code to create a vector that contains the value `2` ten times would be

```{r}
repeating_vector <- rep(2, 10)
repeating_vector
```

To generate a sequence of numbers, such as in the Chicago and Texas spot prices, you may want to use the `seq()` function (see documentation [here](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/seq)). The syntax is `seq(start_value, end_value, by = increment_value)`. The code to create a vector that starts at 1 and increases by the value of 3 until it reaches the value 20 is

```{r}
sequence <- seq(1, 20, by = 3)
sequence
```
You can also use the `seq()` function if you only know the starting value, the increment, and the desired length of the sequence. The code to create a vector that starts at 1 and increases by the value of 3 until the vector reaches the length of 7 is

```{r}
sequence2 <- seq(1, by = 3, length = 7)
sequence2
```
